Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-06 ms.tex
fixed 6.3

If the nuclei of elliptical galaxies are heated in a way similar to that due to the average star formation history of galaxies with
similar mass, then to first order we predict that $\langle
L_{X}/L_{\rm edd}\rangle$ should increase with black hole mass, though
only weakly for $M_{\bullet} \lesssim 3\times 10^{8}M_{\odot}$.  This
result is in tension with the observed down-sizing trend that the
average Eddington ratio is seen to decrease (albeit also weakly)
with $M_{\bullet}$.  However, one must keep in mind the enormous uncertainty in calculating the
luminosity of the accretion flow close to the black hole in a single
waveband based on the feeding rate on larger scales.  If disk outflows carry away most of the infalling mass before it reaches X-ray producing radii (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}), then the
angular momentum of the infalling gas might also influence the X-ray luminosity, in a way
that could depend systematically on the stellar population and hence $M_{\bullet}$.

@@

It is also possible that many of the
\citet{Miller+15} sample of galactic nuclei are not accreting in steady state.  The
value of the X-ray efficiency assumed in the right panel of Figure \ref{fig:miller} represents a maximum value, since it does not account for suppression of the SMBH accretion rate due to outflows from the accretion disk on small
radial scales.  If the actual radiative efficiency is lower, then many of the X-ray detections would lie even further above above the predicted value of $\langle L_X \rangle$ for stable accretion (black line).  Either these
galaxies experience much less heating than average galaxies of the same mass, or they are not accreting stably
at all.  Indeed, for $M_{\bullet} \lesssim
10^{8}M_{\odot}$, $L_{\rm X}$ nearly exceeds that predicted for the
maximum thermally-stable accretion rate, $\dot{M}_{\rm TI}$ (yellow
line).  Since the \citet{Miller+15} sample includes only early-type galaxies with older stellar populations than the average for their black hole mass, it would not be too surprising if many of these galaxies were thermally unstable.

Figure \ref{fig:bh_growth} (top panel) also shows that Compton heating cannot prevent average
accretion rates which are thermally unstable, i.e. $\dot{M}_{\rm C} >
\dot{M}_{\rm TI}$, across all
black hole masses.  When thermal instabilities become manifest, nuclear gas may undergo cyclic oscillations of rapid cooling and high accretion rates, followed by quiescent periods once gas has been consumed and/or
AGN feedback becomes effective.  Such cycles are seen in numerical simulations of elliptical galaxies on 
large scales and long timescales (\citealt{Ciotti+10}), an idea which supported by observations showing that a fraction of early-type galaxies in the local universe have undergone recent ($< 1-2$ Gyr old) star formation episodes (\citealt{Donas+07}).

Similar cyclic AGN activity could occur on the smaller radial scale of
the sphere of influence (e.g.~\citealt{Yuan&Li11},
\citealt{Cuadra+15}).  In this case the large scatter in the X-ray
luminosities shown in Fig \ref{fig:miller} could result from the wide
range of inflow rates experienced over the course of a cyclic episode
between instabilities and periodic nuclear outbursts.  Due to their
deeper potential wells {\bf AG: Cores don't necessarily have deeper
  potential wells overall}, we find that core
galaxies are more likely to possess higher accretion rates and
potentially be thermally unstable for a fixed wind heating rate than
cusp galaxies.  This could help explain why core galaxies show a wide
range of nuclear radio and X-ray luminosities (suggesting cyclic
accretion episodes), while cusp galaxies are confined below a
threshold (e.g., \citealt{Bender+89, Pellegrini99,
  Capetti&Balmaverde05}), consistent with thermally-stable,
steady-state picture. ({\bf AG: This picture may be dubious...Review
  the literature but I think that the higher upper bound for cores is
  just due to a mass selection effect.  BDM: The ghost of Balmaverde
  disagrees.})

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$ in Fig
\ref{fig:miller} indeed represent thermally-unstable galaxies, then a fraction of the upper limits may contribute to a separate thermally stable population, for instance including SgrA*.  However, the possibility of a mixed population of stable versus unstable galaxies calls into question the use of simple extrapolations of power-law fits to the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the occupation fraction of black holes in low mass galaxies (\citealt{Miller+15}).

>>>>

If the nuclei of elliptical galaxies are heated in a way similar to that due to the average star formation history of galaxies with
similar mass, then to first order we predict that $\langle
L_{X}/L_{\rm edd}\rangle$ should be relatively independent of BH mass.  To within the uncertainties, this
result is consistent with the observed down-sizing trend that the average Eddington ratio is seen to decrease (albeit weakly)
with $M_{\bullet}$, although tension is somewhat greater in the $\epsilon_X=
  (\epsilon_{\rm bol}/0.1)(\Mdot/\MdotEdd)$ model.

%However, one must keep in mind the enormous uncertainty in calculating the
%luminosity of the accretion flow close to the black hole in a single
%waveband based on the feeding rate on larger scales.  If disk outflows carry away most of the infalling mass before it reaches X-ray producing radii (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}), then the
%angular momentum of the infalling gas might also influence the X-ray luminosity, in a way
%that could depend systematically on the stellar population and hence $M_{\bullet}$.

@@

It is also possible that many of the
\citet{Miller+15} sample of galactic nuclei are not accreting in steady state.  The
value of the X-ray efficiency assumed in the right panel of Figure \ref{fig:miller} represents a maximum value, since it does not account for suppression of the SMBH accretion rate due to outflows from the accretion disk on small
radial scales (e.g., \citealt{Blandford&Begelman99}).  If the actual radiative efficiency is lower than our fiducial assumption, then many of the X-ray detections would lie above the predicted value of $\langle L_X \rangle$ for stable accretion (black line).  In this case the X-ray detected galaxies either experience less heating than average galaxies of the same mass, or they are not accreting stably at all.  {\bf BDM: we may need to remove - why did this change so much?} Indeed, for $M_{\bullet} \lesssim
10^{8}M_{\odot}$, $L_{\rm X}$ nearly exceeds that predicted for the
maximum thermally-stable accretion rate, $\dot{M}_{\rm TI}$ (yellow
line).  Since the \citet{Miller+15} sample includes only early-type galaxies with older stellar populations than the average for their black hole mass, it would not be too surprising if many of these galaxies were thermally unstable.

Figure \ref{fig:bh_growth} (top panel) also shows that $\dot{M}_{\rm C} >
\dot{M}_{\rm TI}$ across all black hole masses.  Thus, even if Compton heating could act to stabilize the flow (which, as already mentioned, is unlikely), for fiducial parameters it is insufficient to produce a thermally stable accretion rate given conditions corresponding to the average star formation history.  When thermal instabilities become manifest, nuclear gas may undergo cyclic oscillations of rapid cooling and high accretion rates, followed by quiescent periods once gas has been consumed and/or
AGN feedback becomes effective.  Such cycles are seen in numerical simulations of elliptical galaxies on 
large scales and long timescales (\citealt{Ciotti+10}), an idea which is supported by observations showing that a fraction of early-type galaxies in the local universe have undergone recent ($< 1-2$ Gyr old) star formation episodes (\citealt{Donas+07}).

Similar cyclic AGN activity could occur on the smaller radial scale of
the sphere of influence (e.g.~\citealt{Yuan&Li11},
\citealt{Cuadra+15}).  In this case the large scatter in the X-ray
luminosities shown in Fig \ref{fig:miller} could result from the wide
range of inflow rates experienced over the course of a cyclic episode
between instabilities and periodic nuclear outbursts.  Due to their
deeper potential well on scales of the sphere of influence, core galaxies are more likely to possess higher accretion rates and
potentially be thermally unstable for a fixed wind heating rate than
cusp galaxies.

%This could help explain why core galaxies show a wide
%range of nuclear radio luminosities (suggesting cyclic
%accretion episodes), while cusp galaxies are confined below a
%threshold (e.g., \citealt{Bender+89, Pellegrini99,
 % Capetti&Balmaverde05}), consistent with thermally-stable,
%steady-state picture. ({\bf AG: This picture may be dubious...Review
 % the literature but I think that the higher upper bound for cores is
  %just due to a mass selection effect.  BDM: Balmaverde
 % disagrees.})

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$ in Fig
\ref{fig:miller} are indeed in a state of thermally-unstable cyclic behavior, then a subset of the galaxies with X-ray upper limits may contribute to a separate steady-state, thermally stable population, for instance including SgrA* (which has a much lower X-ray luminosity for its black hole mass than predicted by the \citet{Miller+15} trend).  One consequence of a potentially bimodal population of steady and cyclic galactic nuclei calls into question the physical justification of using simple extrapolations of power-law fits to the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the occupation fraction of SMBHs in the nuclei of low mass galaxies.
